Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (*Sida fallax*)

### 2023-10-26 ms/ms.qmd
revising ms

Amphistomatous leaves should be common because, all else being equal, a leaf with a given number of stomata per area could increase its photosynthetic rate simply by apportioning approximately half its stomata to each surface [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984; @gutschick_photosynthesis_1984-1]. The key difference between a hypo- and amphistomatous leaf, holding all other factors constant, is that an amphistomatous leaf has two parallel diffusion paths through the internal airspace to any given chloroplast. Those airspaces pose a resistance for CO$_2$ diffusion, so CO$_2$ concentration drops as it approaches chloroplasts. Shorter pathways mean a smaller drop in CO$_2$ concentration. Thus, chloroplasts in amphistomatous leaves experience higher CO$_2$ concentrations than in hypostomatous leaves, thereby increasing photosynthesis. The airspace resistance (or its inverse, the airspace conductance, $g_\mathrm{ias}$) is rarely measured directly and there is disagreement between empirical [@parkhurst_intercellular_1990; @morison_lateral_2005; @evans_resistances_2009; @tomas_importance_2013; @earles_beyond_2018; @santrucek_partitioning_2019; @nobel_physicochemical_2020; @harwood_understanding_2021; @marquez_assessing_2023] and theoretical models [@tholen_mechanistic_2011; @ho_three-dimensional_2016; @theroux-rancourt_maximum_2021]. The $g_\mathrm{ias}$ in thin, porous leaves may be so large as to be inconsequential given much lower conductances for other components of the diffusion pathway, whereas the $g_\mathrm{ias}$ of thick leaves with little airspace may greatly hinder CO$_2$ diffusion to chloroplasts. Amphistomy should confer the largest photosynthetic benefit in leaves with intrinsically low $g_\mathrm{ias}$.

@@

The association between high light and amphistomy suggests that 'sun' leaves have the most to gain in terms of increased photosynthesis from having stomata on both surfaces, as @mott_adaptive_1982 hypothesized. @parkhurst_adaptive_1978 proposed quantifying this benefit as 'amphistomy advantage' ($\mathrm{AA}$), which we adopt here with some modification (see Materials and Methods). This hypothesis has never been tested directly by comparing the photosynthetic rate of an amphistomatous leaf to that of an otherwise identical hypostomatous leaf with the same total stomatal conductance under the same conditions. We propose a straightforward method to do this by experimentally creating a pseudohypostomatous leaf with gas exchange blocked through the upper surface (see Materials and Methods). One reason that sun leaves might have greater $\mathrm{AA}$ is that they are usually thicker or denser [@poorter_metaanalysis_2019], resulting in lower $g_\mathrm{ias}$ [@parkhurst_adaptive_1978]. A nonmutually exclusive hypothesis is that if sun leaves have a stomatal density ratio closer to 0.5 (same density on each leaf surface), this will confer a greater advantage than an amphistomatous leaf with most stomata on one surface. In other words, amphistomy doesn't make much difference if one leaf surface has few open stomata on it. We therefore predict that sun leaves will have greater $\mathrm{AA}$ possibly because they have thicker leaves and/or $\mathrm{SR}$ closer to 0.5. We actually report $g_{\mathrm{smax,ratio}}$, which is similar to $\mathrm{SR}$ except that it accounts for differences in both stomatal density and size between surfaces.

@@

We estimated the stomatal density and size on ab- and adaxial leaf surfaces from all leaves. For pubescent leaves (usually coastal), we dried and pressed leaves for $\approx 1$ week [@hill_temperature_2014], carefully scraped trichomes off with a razor blade, and rehydrated the leaf. For glabrous leaves, we used fresh leaves. We applied clear nail polish to both leaf surfaces of fresh or rehydrated leaves in the middle of the lamina away from major veins. After nail polish dried, we mounted impressions on a microscope slide using transparent tape [@mott_amphistomy_1991]. We digitized a portion of each leaf surface impression using a brightfield microscope (Leica DM2000, Wetzlar, Germany). We counted all stomata and divided by the visible leaf area (0.890 mm$^2$) to estimate density and measured guard cell length from five randomly chosen stomata per field using ImageJ [@schneider_nih_2012].

@@

The log-response ratio is straightforward to interpret because values above 0 indicate a photosynthetic advantage of amphistomy, whereas values less than 0 indicate a disadvantage. The response ratio is also better than the absolute difference because it indicates a proportional change in $A$, which facilitates comparisons across leaves and environments with different baseline photosynthetic rates. The irradiance of the light source in the pseudohypo leaf is higher because the propafilm reduces transmission. To compensate for reduced transmission, we increased incident $\mathrm{PPFD}$ for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm. We also set the stomatal conductance ratio, for purposes of calculating boundary layer conductance, to 0 for pseudohypo leaves following manufacturer directions.

>>>>

Amphistomatous leaves should be common because, all else being equal, a leaf with a given number of stomata per area could increase its photosynthetic rate simply by apportioning approximately half its stomata to each surface [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984; @gutschick_photosynthesis_1984-1]. The key difference between a hypo- and amphistomatous leaf, holding all other factors constant, is that an amphistomatous leaf has two parallel diffusion paths through the internal airspace to any given chloroplast. Those airspaces pose a resistance for CO$_2$ diffusion, so CO$_2$ concentration drops as it approaches chloroplasts. Shorter pathways mean a smaller drop in CO$_2$ concentration. Thus, chloroplasts in amphistomatous leaves experience higher CO$_2$ concentrations than in hypostomatous leaves, thereby increasing photosynthesis. The airspace resistance (or its inverse, the airspace conductance, $g_\mathrm{ias}$) is rarely measured directly and there is disagreement between empirical [@parkhurst_intercellular_1990; @morison_lateral_2005; @evans_resistances_2009; @tomas_importance_2013; @earles_beyond_2018; @santrucek_partitioning_2019; @nobel_physicochemical_2020; @harwood_understanding_2021; @marquez_assessing_2023] and theoretical models [@tholen_mechanistic_2011; @ho_three-dimensional_2016; @theroux-rancourt_maximum_2021]. The $g_\mathrm{ias}$ in thin, porous leaves may be so large as to be inconsequential given much lower conductances for other components of the diffusion pathway, whereas the $g_\mathrm{ias}$ of thick leaves with little airspace may greatly hinder CO$_2$ diffusion to chloroplasts. Amphistomy should confer the largest photosynthetic benefit in leaves with intrinsically low $g_\mathrm{ias}$. The airspace conductance is one component of the overall mesophyll conductance, $g_\mathrm{m}$, which is often strongly influenced by the chloroplast surface area exposed to airspace and mesophyll cell wall thickness [@evans_resistances_2009; @gago_photosynthesis_2020; @flexas_cell_2021]. Hence, thicker leaves may compensate for lower $g_\mathrm{ias}$ through increased chloroplast surface area exposed to airspace [@terashima_irradiance_2006], but would still benefit from amphistomy.

@@

The association between high light and amphistomy suggests that 'sun' leaves have the most to gain in terms of increased photosynthesis from having stomata on both surfaces, as @mott_adaptive_1982 hypothesized. @parkhurst_adaptive_1978 proposed quantifying this benefit as 'amphistomy advantage' ($\mathrm{AA}$), which we adopt here with some modification (see Materials and Methods). This hypothesis has never been tested directly by comparing the photosynthetic rate of an amphistomatous leaf to that of an otherwise identical hypostomatous leaf with the same total stomatal conductance under the same conditions. We propose a straightforward method to do this by experimentally creating a pseudohypostomatous leaf with gas exchange blocked through the upper surface (see Materials and Methods). One reason that sun leaves might have greater $\mathrm{AA}$ is that they are usually thicker or denser [@poorter_metaanalysis_2019], which will often result in lower $g_\mathrm{ias}$ either by increasing the diffusion path length [@parkhurst_adaptive_1978] or making the airspace less porous. A nonmutually exclusive hypothesis is that if sun leaves have a stomatal density ratio closer to 0.5 (same density on each leaf surface), this will confer a greater advantage than an amphistomatous leaf with most stomata on one surface. In other words, amphistomy doesn't make much difference if one leaf surface has few open stomata on it. We therefore predict that sun leaves will have greater $\mathrm{AA}$ possibly because they have thicker leaves and/or $\mathrm{SR}$ closer to 0.5. We actually report $g_{\mathrm{smax,ratio}}$, which is similar to $\mathrm{SR}$ except that it accounts for differences in both stomatal density and size between surfaces.

@@

We estimated the stomatal density and size on ab- and adaxial leaf surfaces from all leaves. For pubescent leaves (usually coastal), we dried and pressed leaves for $\approx 1$ week [@hill_temperature_2014], carefully scraped trichomes off with a razor blade, and rehydrated the leaf. Rehydration restores leaf area to its fresh value [@blonder_leaf-area_2012]. For glabrous leaves, we used fresh leaves. We applied clear nail polish to both leaf surfaces of fresh or rehydrated leaves in the middle of the lamina away from major veins. After nail polish dried, we mounted impressions on a microscope slide using transparent tape [@mott_amphistomy_1991]. We digitized a portion of each leaf surface impression using a brightfield microscope (Leica DM2000, Wetzlar, Germany). We counted all stomata and divided by the visible leaf area (0.890 mm$^2$) to estimate density and measured guard cell length from five randomly chosen stomata per field using ImageJ [@schneider_nih_2012].

@@

The log-response ratio is commonly used social and biological sciences (e.g. @hedges_metaanalysis_1999). It is straightforward to interpret because values above 0 indicate a photosynthetic advantage of amphistomy, whereas values less than 0 indicate a disadvantage. The response ratio is also better than the absolute difference because it indicates a proportional change in $A$, which facilitates comparisons across leaves and environments with different baseline photosynthetic rates. The irradiance of the light source in the pseudohypo leaf is higher because the propafilm reduces transmission. To compensate for reduced transmission, we increased incident $\mathrm{PPFD}$ for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm. We also set the stomatal conductance ratio, for purposes of calculating boundary layer conductance, to 0 for pseudohypo leaves following manufacturer directions.
